---
title: "Activity Monitoring Data"
author: "zrseah"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


## Loading and preprocessing data 

### Import Libraries 
```{r results='hide'}
library('magrittr')
library('dplyr')
library('ggplot2')
library('tidyr')
library('patchwork')
```

### Load Data
```{r}
activity <- read.csv("~/Online Courses/Johns Hopkins University Data Science Foundations using R Specialization/Assignment/Module 5/RepData_PeerAssessment1/activity.csv")
```

### View structure of data
```{r}
str(activity)
```

### Data Cleaning
```{r}
# Convert date into class date 
activity <- activity %>%  
  mutate(date = as.Date(as.character(date), "%Y-%m-%d"))

# Check class of date 
class(activity$date)
```


## Analysis

### Total number of steps taken per day
```{r}
# Create a data frame that shows the total number of steps each day 
dailysteps <- aggregate(steps ~ date, data = activity, sum, na.action = na.pass)
colnames(dailysteps)[colnames(dailysteps) == 'steps'] <- 'total_steps'

# Plot histogram of total steps taken each day
ggplot(dailysteps, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "lightsteelblue4") +
  labs(x = "Total_Steps", y = "Frequency", title = "Total Number of Steps Taken Each Day")

```

### Mean and median total number of steps taken per day
```{r}
tdsmean <- round(mean(dailysteps$total_steps, na.rm = TRUE), 2)
tdsmedian <- round(median(dailysteps$total_steps, na.rm = TRUE), 2)
tdsmean_median <- data.frame(
  Metric = c("Mean", "Median"),
  Steps = c(tdsmean, tdsmedian)
)
tdsmean_median
```

### Average daily activity pattern
```{r}
# Create a data frame on average number of steps taken for all days per interval 
intervalsteps <- aggregate(steps ~ interval, data = activity, mean)
intervalsteps$steps <- round(intervalsteps$steps, 2)
colnames(intervalsteps)[colnames(intervalsteps) == "steps"] <- "average_steps"

# Plot average daily activity pattern
ggplot(intervalsteps, aes(x = interval, y = average_steps)) +
  geom_line(type = "l", color = "lightsteelblue4") +
  labs(x = "5_Minute_Interval", y = "Average_Steps", title = "Average Daily Activity Pattern")
```




Interval 835 contains the maximum number of steps.




```{r}
# Interval with the maximum number of steps 
## Find index of interval with max average steps 
which.max(intervalsteps$average_steps)
## Find the interval through index 
intervalsteps$interval[104]
```

### Imputing Missing Values 




Total number of missing values is 2304.




```{r}
sum(is.na(activity))
```




Missing values were filled up using mean values from 5-minute interval.




### New dataset with missing data filled in
```{r}
# Merge mean steps per 5-minute interval with original data frame 
activity2 <- activity %>%  
  left_join(intervalsteps, by = "interval")

# Fill missing value with mean steps per 5-minute interval
activity2 <- activity2 %>% 
  mutate(steps = ifelse(is.na(steps), average_steps, steps)) %>% 
  select(-average_steps)

# Rearrange column 
activity2 <- activity2[, c(2, 3, 1)]
head(activity2)
```

### Total number of steps taken per day
```{r}
# Create a data frame that shows the total number of steps each day 
dailysteps2 <- aggregate(steps ~ date, data = activity2, sum)
colnames(dailysteps2)[colnames(dailysteps2) == "steps"] <- "total_steps"

# Plot histogram of total steps taken each day 
ggplot(dailysteps2, aes(x = total_steps)) + 
  geom_histogram(binwidth = 1000, fill = "lightsteelblue4") +
  labs(x = "Total_Steps", y = "Frequency", title = "Total Number of Steps Taken Each Day")
```

### Mean and median total number of steps taken per day
```{r}
tdsmean2 <- round(mean(dailysteps2$total_steps), 2)
tdsmedian2 <- round(median(dailysteps2$total_steps), 2)
tdsmean_median2 <- data.frame(
  Metric = c("Mean", "Median"),
  Steps = c(tdsmean2, tdsmedian2)
)
tdsmean_median2
```




Mean and median values are rather close to the values from first part of the assignment. 




### Differences in activity patterns between weekdays and weekdends

#### Create new factor variable in data set 
```{r}
day <- weekdays(activity2$date)
activity3 <- activity2 %>% 
  mutate(day) 
activity3  <- activity3 %>%
  mutate(day_type = ifelse(day %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
```




Average steps taken are higher during the weekends as compared to weekdays. 




#### Average daily activity pattern
```{r}
# Create a subset data for average number of steps per interval for weekdays 
intervalstepsweekdays <- subset(activity3, day_type == "Weekday") 
intervalstepsweekdays <- aggregate(steps ~ interval, data = intervalstepsweekdays, sum)

# Create a subset data for average number of steps per interval for weekends
intervalstepsweekends <- subset(activity3, day_type == "Weekend") 
intervalstepsweekends <- aggregate(steps ~ interval, data = intervalstepsweekends, sum)

# Plot average daily activity pattern for weekdays and weekends 
p1 <- ggplot(intervalstepsweekdays, aes(x = interval, y = steps)) + 
        geom_line(type = "l", color = "lightsteelblue4") +
        labs(x = "5_Minute_Interval", y = "Average_Steps", title = "Weekdays")

p2 <- ggplot(intervalstepsweekends, aes(x = interval, y = steps)) + 
        geom_line(type = "l", color = "lightsteelblue4") +
        labs(x = "5_Minute_Interval", y = "Average_Steps", title = "Weekends")

combinedplot <- p1 + p2 + plot_layout(nrow = 2) +
  plot_annotation(title = "Average Daily Activity Pattern")

combinedplot
```